---
title: "Assignment: Course Project 1"
author: "Andrea Eoli"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.path = "./figures/")

library(dplyr)
library(lubridate)
library(tidyverse)
library(ggplot2)

```

## Loading and preprocessing the data
```{r load}
df <- tibble(read.csv(unzip("activity.zip")))
str(df)
df$date <- as.Date.character(df$date)
```


## What is mean total number of steps taken per day?
```{r q1}
q1 <- df %>% drop_na() %>% group_by(date) %>% summarize(tot_steps = sum(steps))

ggplot(q1, aes(x=date, y=tot_steps)) + 
  geom_bar(stat = "identity", fill = "#69b3a2") + ggtitle("Total steps by day")

mean <- mean(q1$tot_steps)
median <- median(q1$tot_steps)

```

The mean of the total steps taken per day is `r as.integer(mean)`, while the median is `r as.integer(median)`.


## What is the average daily activity pattern?
```{r q2}
q2 <- df %>% drop_na() %>% group_by(interval) %>% summarize(avg_steps = mean(steps))

plot(q2, type = "l", main = "Average steps by 5-min interval")

max <- as.numeric(q2[q2$avg_steps == max(q2$avg_steps),"interval"])
```

The 5-minute interval, on average across all the days in the dataset, that contains the maximum number of steps is `r max`.


## Imputing missing values
```{r q3}
# Count NAs
NAs <- sum(is.na(df))

# Impute NAs with AVG(interval)
q3 <- df
q3$steps <- ave(q3$steps, q3$interval, FUN=function(x) 
  ifelse(is.na(x), mean(x, na.rm = TRUE), x))

# Plot
q3_grouped <- q3 %>% group_by(date) %>% summarize(tot_steps = sum(steps)) 

q3_grouped %>% 
  ggplot(aes(x=date, y=tot_steps)) + 
  geom_bar(stat = "identity", fill = "#69b3a2") + ggtitle("Total steps by day")

mean2 <- mean(q3_grouped$tot_steps)
median2 <- median(q3_grouped$tot_steps)

if (mean == mean2 & median == median2) {q3_comp <- "both are identical to the previously calculated values"
} else{q3_comp <- "there are some differences compared to the previous values"}

```

The total number of missing values in the dataset is `r NAs`. The new mean is `r as.integer(mean2)` and the new median is `r as.integer(median2)`, `r q3_comp`. When comparing the imputed dataset with the previous one, the main difference is that there are `r nrow(q3_grouped) - nrow(q1)` new observations/rows.

## Are there differences in activity patterns between weekdays and weekends?
```{r q4}
q4 <- q3

q4$weekday <- weekdays(q4$date)
q4$type <- factor(ifelse(q4$weekday %in% c("Saturday","Sunday"), "weekend","weekday"))


q4_avg <- q4 %>% group_by(type,interval) %>% summarize(avg_steps = mean(steps))

ggplot(data = q4_avg, aes(interval, avg_steps)) +
  geom_line(color = "steelblue", size = 1) +
  labs(title = "Average number of steps taken by 5-min interval",
       y = "Average steps", x = "Intervals") + 
  facet_wrap(~ type)

```
